Evaluating FAIR Digital Object as a distributed object system

### 2023-05-18 latex/main.tex
more on EOSC IF

In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.

@@

\textbf{PID Profiles \& Attributes} \autocite{fdo-PIDProfileAttributes} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile. This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group \autocite{weigelRDARecommendationPID2018}. This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}\autocite{islam_2023}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.

@@

\textbf{Upload of FDO} \autocite{fdo-FDO-Upload} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve). ResourceSync \autocite{ResourceSyncFrameworkSpecification} is suggested as one type of service to list FDOs. This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).

@@

It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.

@@

The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee \autocite{berners-leeWeavingWebOriginal1999}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs. The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}. The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.

@@

While triple-based knowledge representations were not new \autocite{stanczykProcessModellingInformation1987}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \autocite{rfc3986} are generalised forms of URLs that include locator-less identifiers such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years \autocite{InfoURIRegistry}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \autocite{jutyIdentifiersOrgMIRIAM2011}. RDF 1.1 formats use Unicode to support IRIs \autocite{rfc3987}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \autocite{sauermannCoolURIsSemantic2011}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.

The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data). Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term. The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource. By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs. The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.

The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is). More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \autocite{horrocksSemanticWebISWC2002} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.

@@

The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources. This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \autocite{grothAPIcentricLinkedData2014b}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \autocite{kamdarSystematicAnalysisTerm2017a}. Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \autocite{kleinScholarlyContextNot2014a}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \autocite{polleresMoreDecentralizedVision2020a}. Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \autocite{carrieroLandscapeOntologyReuse2020a}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \autocite{w3-rdf-schema}, OWL \autocite{w3-owl2-overview}, OBO \autocite{tirmiziMappingOBOOWL2011a}, SKOS \autocite{w3-skos-primer}), hash vs slash, HTTP status codes and PID redirection strategies \autocite{sauermannCoolURIsSemantic2011}.

@@

Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data. In addition to several such embedded \emph{microformats} (Open Graph \autocite{OpenGraphProtocol}, RDFa \autocite{w3-rdfa-primer}, Microdata \autocite{HTMLStandard}) we find JSON-LD \autocite{w3-json-ld} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications. JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 44.4\% of the top 10 million websites \autocite{UsageStatisticsJSONLD}.

Recently there has been a renewed emphasis to improve the \emph{Developer Experience} \autocite{DesigningLinkedData2018} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \autocite{w3-shacl} or ShEx \autocite{ShapeExpressionsShEx}) can be used to validate RDF Data \autocite{gayoValidatingRDFData2017a,thorntonUsingShapeExpressions2019a} before consuming it programmatically, or reshaping data to fit other models. While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \autocite{klimekSurveyToolsLinked2019a}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.

@@

\subsection*{Comparing FDO and existing approaches}\label{sec:comparing}

@@

\begin{enumerate}
\tightlist
\item
  An Interoperability Framework and Distributed Platform for Fast Data Applications \autocite{delgadoInteroperabilityFrameworkDistributed2016a}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
\item
  The FAIR Digital Object guidelines \autocite{boninoFAIRDigitalObject}, validated against its implementations for completeness.
\item
  A Comparison Framework for Middleware Infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
\item
  Cross-checks against RDA's FAIR Data Maturity Model \autocite{bahimFAIRDataMaturity2020a} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
\item
  EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
\end{enumerate}

The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (\vpageref*{sec:next-step-fdo}).

Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices. For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data in general.

\section*{Methods}
\subsection*{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}

@@

Based on the analysis shown in Table \vref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:

The Web has already showed us how one can compose workflows of hetereogeneous Web Services \autocite{wolstencroftTavernaWorkflowSuite2013d}. However, this is mostly done via developer or human interaction \autocite{lamprechtPerspectivesAutomatedComposition2021b}. Similiarly, FDO does not enable automatic composition because operation semantics are not well defined. There is a question as to whether the plethora of documentation and broad developer usage that is available for Web APIs can be developed for FDO.

A difference between Web technologies and FDO is the stringency of the requirements for both syntax and semantics. Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON. On the semantic front, FDO requires that every object have a well-defined type and structured form. This is clearly not the case on the Web.
In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed. These have shown interoperability. The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles \autocite{ModernStandardsParadigm} to embrace openness, transparency, and broad consensus. In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \autocite{looFirstInternationalConference2022}. This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP \autocite{DOIPAPIHTTPa}.

@@

\textbf{TODO}: Introduce EOSC IF

The EOSC Interoperability Framework \autocite[][section 3.6]{corchoEOSCInteroperabilityFramework2021b} recommends:

Observations:

\begin{itemize}
\tightlist
\item
  The recommendations from EOSC IF are at a higher level that mainly affect governance and practices by communities
\item
  Technical aspects highlighted by EOSC IF
\item
  Search/indexing is important FAIR aspect for Findability, but is poorly supported by current FDO and Linked Data. There is a strong role for organizations like EOSC to provide broader registries than more specialised metadata federations like OpenAIRE.
\item
  FDO principles have strong recommendations for community development of organisational aspects.
\item
  Both FDO and LD are weak on legal aspects like licensing, privacy and usage policies -- these are essential for cross-institutional and cross-repository access of FAIR objects
\end{itemize}

>>>>

In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. We do so by looking through the lens of development of these technologies over time, including future directions.

@@

\textbf{PID Profiles \& Attributes} \autocite{fdo-PIDProfileAttributes} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile. This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group \autocite{weigelRDARecommendationPID2018}. This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record} \autocite{islam_2023}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.

@@

\textbf{Upload of FDO} \autocite{fdo-FDO-Upload} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve). ResourceSync \autocite{ResourceSyncFrameworkSpecification} is suggested as one type of service to list FDOs. This document highlights potential practices by repositories and their clients, without adding any particular requirements.

@@

It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.

@@

The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee \autocite{berners-leeWeavingWebOriginal1999}, at a time that the Web had already become widely established for information exchange, being a global set of hypermedia documents which are cross-related using universal links in the form of URLs. The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}. The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.

@@

While triple-based knowledge representations were not new \autocite{stanczykProcessModellingInformation1987}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \autocite{rfc3986} are generalised forms of URLs that include locator-less identifiers such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years \autocite{InfoURIRegistry}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \autocite{jutyIdentifiersOrgMIRIAM2011}. RDF 1.1 formats use Unicode to support IRIs \autocite{rfc3987}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \autocite{sauermannCoolURIsSemantic2011}.}, the Semantic Web builds a self-described system of types and properties, where the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}. By applying these principles as well to any kind of resource that could be described at a URL, this then forms a global distributed Semantic Web.

The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data). Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term. The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) with a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource. By using \emph{content negotiation} in HTTP\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}}, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs. The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.

The mid-2000s saw large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL \autocite{w3-owl2-overview}, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is). More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \autocite{horrocksSemanticWebISWC2002} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.

@@

The move towards \emph{Open Science} data sharing practices did from the late 2000s encourage knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources. This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \autocite{grothAPIcentricLinkedData2014b}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \autocite{kamdarSystematicAnalysisTerm2017a}. Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \autocite{kleinScholarlyContextNot2014a}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \autocite{polleresMoreDecentralizedVision2020a}. Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \autocite{carrieroLandscapeOntologyReuse2020a}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \autocite{w3-rdf-schema}, OWL \autocite{w3-owl2-overview}, OBO \autocite{tirmiziMappingOBOOWL2011a}, SKOS \autocite{w3-skos-primer}), hash vs slash, HTTP status codes and PID redirection strategies \autocite{sauermannCoolURIsSemantic2011}.

@@

Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data. In addition to several such embedded \emph{microformats} \autocite{OpenGraphProtocol,w3-rdfa-primer,HTMLStandard} or we find JSON-LD \autocite{w3-json-ld} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications. JSON-LD is as of 2023-05-18 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 45\% of the top 10 million websites \autocite{UsageStatisticsJSONLD}.

Recently there has been a renewed emphasis to improve the \emph{Developer Experience} \autocite{DesigningLinkedData2018} for consumption of Linked Data, for instance RDF Shapes -- expressed in SHACL \autocite{w3-shacl} or ShEx \autocite{ShapeExpressionsShEx} -- can be used to validate RDF Data \autocite{gayoValidatingRDFData2017a,thorntonUsingShapeExpressions2019a} before consuming it programmatically, or reshaping data to fit other models. While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying Semantic Web technologies, which hampers adaption by non-LD experts \autocite{klimekSurveyToolsLinked2019a}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.

@@

Our main motivation for this article is to investigate how the promises of FAIR Digital Objects may differ from the learnt experiences of Linked Data and the Web. We also reflect back from FDO's motivation of machine-actionability to consider the Web as a distributed computational system.

@@

\begin{enumerate}
\tightlist
\item
  An Interoperability Framework and Distributed Platform for Fast Data Applications \autocite{delgadoInteroperabilityFrameworkDistributed2016a}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view.
\item
  The FAIR Digital Object guidelines \autocite{boninoFAIRDigitalObject}, validated against its current implementations for completeness.
\item
  A Comparison Framework for Middleware Infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods.
\item
  Cross-checks against RDA's FAIR Data Maturity Model \autocite{bahimFAIRDataMaturity2020a} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness.
\item
  EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective.
\end{enumerate}

The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref*{sec:next-step-fdo}).

Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices. For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data in general practice.

\subsection*{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}

@@

Based on the analysis shown in Table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:

The Web has already showed us how one can compose workflows of hetereogeneous Web Services \autocite{wolstencroftTavernaWorkflowSuite2013d}. However, this is mostly done via developer or human interaction \autocite{lamprechtPerspectivesAutomatedComposition2021b}. Similiarly, FDO does not enable automatic composition because operation semantics are not well defined. There is a question as to whether the extebsuve documentation and broad developer usage that is available for Web APIs could potentially be utillised for FDO.

A difference between Web technologies and FDO is the stringency of the requirements for both syntax and semantics. Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON. On the semantic front, FDO mandates that every object have a well-defined type and structured form. This is clearly not the case on the Web.
In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed. These have shown interoperability. The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles \autocite{ModernStandardsParadigm} to embrace openness, transparency, and broad consensus. In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first international FDO conference \autocite{looFirstInternationalConference2022}. This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP \autocite{DOIPAPIHTTPa}.

@@

The European Open Science Cloud (EOSC) is a large EU initiative to promote Open Science by implementing a joint research infrastructure by federating existing and new services and focusing on interoperability, accessability, best practices as well as technical infrastructure \autocite{10.2777/940154}. The EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} details the principles for creating a common way to achieve interoperability between all digital aspects of research activities in EOSC, including data, protocols and software. The recommendations are realized through 4 layers, Technical (e.g. protocols), Semantic (e.g. metadata models), Organisational (e.g. recommendations) and Legal (e.g. agreements), with a particular aim to address the FAIR interoperability principles and building on the concept of FAIR Digital Objects.

In Table \vref{tbl:eosc} we review the EOSC Interoperability Framework (EOSC IF) recommendations, and evaluate to what extent they are addressed by the principles of FDO and Linked Data or their common implementations.

Firstly, we observe that the EOSC IF recommendations are at a high level, mainly affecting governance and practices by communities. This \emph{Organizational} level is also highlighted by the FDO recommendations, for instance the FDO Typing \autocite{fdo-TypingFDOs} propose a governance structure to recognize community-endorsed services. While these community aspects are not mandated by Linked Data practices, best practices have become established for aspects like ontology development \autocite{10.1186/s13326-021-00240-6}. EOSC IF's technical layer is likewise at a architecturally high level, such as service-level agreements, but also highlight PID policies which is strongly required by FDO, while Linked Data communities choose PID practices separately. The recommendations for the Semantic layer, is largely already implemented by Linked Data practices, yet for FDO mostly consist of encouragements. For instance \emph{clear definitions of semantic concepts} is required by FDO guidelines, but how to technically define them has not been formalised by FDO specifications.

The Legal layer of interoperability is perhaps the one most emphasised by EOSC, by enabling collaboration across organizational barriers to joinly build a research infrastructure, but this is an area that both FDO and Linked Data are relatively weak in directly supporting. The EOSC IF recommendations in this layer are still largely related to governance practices and metadata, for instance licensing, privacy and usage policies; yet these are essential for cross-institutional and cross-repository access of FAIR objects. Likewise, search and indexing is important FAIR aspect for Findability, but is poorly supported globally by FDO and Linked Data. Efforts such as Open Research Knowledge Graph (ORKG) \autocite{10.1007/978-3-030-30760-8_31}, DataCite's PID Graph \autocite{10.5438/jwvf-8a66} and Google Knowledge Graph \autocite{singhal2012} have improved programmatic findability to some degree, however not significantly for domain-specific semantic artefacts, currently scattered across multiple semantic catalogues \autocite{10.48550/arXiv.2305.06746}.  There is a strong role for organizations like EOSC to provide such broader registries, moving beyond scholarly output metadata federations. The EOSC Marketplace\footnote{\url{https://marketplace.eosc-portal.eu/}} has for instance recently been expanded to include training material, software and data sources.
